Estimating the number of clusters using Cross Validation

### 2015-06-23 cvclust.tex
revise self-consistency lemma 1

To choose the folds, we randomly partition the rows and columns into $K$ and
$L$ subsets, respectively.  Each fold is indexed by a pair $(r,s)$ of
integers, with $r \in \{1, \dotsc, K\}$ and $s \in \{1, \dotsc, L\}$.  Fold
$(r,s)$ treats the $r$th row subset as ``test'', and the $s$th column subset
as ``response''.  Following \citet{owen2009bi}, we take $K = L = 2$.  For the
number of clusters, we select the value of $k$ that minimizes the average of
$\CV(k)$ over all $K \times L$ folds.

@@

It will suffice to analyze a single fold of the cross-validation procedure.
As in in section~\ref{sec:gabriel-cv-algorithm} we assume that the $P$
variables of the data set have been partitioned into $p$ predictor variables
represented in vector~$X$ and $q$ response variables represented in
vector~$Y$.  The $N$ observations have been divided into two sets: $n$ train
observations and $m$ test observations.

We make the following assumptions:
\begin{enumerate}[label={\arabic*.},leftmargin=*]

  \item There are $k$ true centers $\mu(1), \dotsc,\mu(k)$, with the $g$th
    cluster center partitioned as $\mu(g) = \bigl(\muX(g), \muY(g)\bigr)$ with
    $\muX(g) \in \R^p$ and $\muY(g) \in \R^q$ for $g = 1, \dotsc, k$.

  \item The vectors $\muX(1), \dotsc,\muX(k)$ are all distinct; likewise,
    $\muY(1), \dotsc,\muY(k)$ are all distinct.

  \item The training set contains at least one member of each cluster.

  \item The test set contains at least one member of each cluster.

  \item Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, k \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.

@@

\begin{theorem}
Given assumptions $1-5$, $ E\left(Error_k\right) > 0$ for $k<K$ and  $ E\left(Error_k\right) = 0$ for $k=K$. Hence if $k \leq K$, the Gabriel algorithm selects $k=K$ with probability $1$.
\end{theorem}
\noindent
Not that $k > K$ is not considered here because there are only $K$ unique points in $\mathbf{X}$. K-means will not work for parameter which is larger than the number of unique points in data $\mathbf{X}$.\\\\ 
\noindent
The proof of above theorem follows from following two lemmas:
\begin{lemma}
	If the assumptions of Theorem $1$ hold, then when $k<K$, the CV Error $>0$
\end{lemma}
\begin{proof}
	It is sufficient to prove that when $k<K$,  $Error_k>0$. By the definition of $Error_k$,
\begin{equation}
Error_k = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2
\end{equation}
	where $\hat{\mu}^R \left(g_j\right)$ is the center of cluster $g_j$ returned from K-means. By assumption $4$, test set contains at least one member of each cluster. Since there are $K$ true centers, and by assumption $5$, $x_i = \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\} $, test set contains $K$ centers. It means there are $K$ distinct $x^R_i \left(=\mu^R \left(c_i\right)\right)$ in $\mathbf{X}^R_{test}$ set by assumption $2$. However, in equation $(1)$, there are only $k$ distinct $\hat{\mu}^R \left(g_j\right)$.  Because $k < K$, it means not every element $||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 = 0$ in equation $(1)$. Therefore,  $$Error_k = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 > 0$$
\end{proof}
\begin{lemma}
	If the assumptions of Theorem $1$ hold, when $k=K$, the CV Error $=0$
\end{lemma}
\begin{proof}
	It's sufficient to show that when $k=K$, $Error_k=0$. By assumption $3$ and $5$, training set contains $K$ centers. Which also implies $\mathbf{X}^R_{train}$ contains $K$ distinct $x^R_i$ $(\mu^R_1,...,\mu^R_K)$. When apply K-means on $\mathbf{X}^R_{train}$ with parameter $K$, the unique solution of 
	\[ \underset{\left(\hat{\mu}^R\left(g_1\right),...,\hat{\mu}^R\left(g_K\right)\right)}{\operatorname{argmin}} \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^R_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \]
	is 
\begin{equation}	
	\{\hat{\mu}^R \left(g_j\right) = \mu^R_j, j=1,...,K\}
\end{equation}	
with 
\begin{equation}	 
	\hat{g} \left(x^R_i\right) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
\end{equation}	
Where $\hat{g} \left(x^R_i\right) \in  \mathcal{G}$ is the assigned cluster of $x^R_i$ by K-means. One can easily check that such solution is unique and above objective function attains its lower bound $0$.\\\\
By Proposition $1$, $\mathcal{C}$ has re-substitution error $0$, which means its prediction error is $0$ on $\mathbf{X}^L_{test}$ as $\forall x^L_i \in \mathbf{X}^L_{test}, x^L_i \in \mathbf{X}^L_{train}$ by assumption $1,3,5$. Hence the predicted cluster $\hat{g} \left(x^L_i\right)$ for $x_i$ is the correct one $\hat{g} \left(x^R_i\right)$, which determined by $x^R_i$. By equation $(3)$, we have $x^R_i = \mu^R \left(c_i\right) = \mu^R_j$ if $\hat{g} \left(x^L_i\right) = \hat{g} \left(x^R_i\right) = g_j$. Therefore,
\[ Error_K = \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 =  \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - \mu^R_j||^2= 0\]
The last equation holds because of equation $(2)$.
\end{proof}

>>>>

To choose the folds, we randomly partition the rows and columns into $K$ and
$L$ subsets, respectively.  Each fold is indexed by a pair $(r,s)$ of
integers, with $r \in \{1, \dotsc, K\}$ and $s \in \{1, \dotsc, L\}$.  Fold
$(r,s)$ treats the $r$th row subset as ``test'', and the $s$th column subset
as ``response''.  We typically take $K = 5$ and $L = 2$.  For the number of
clusters, we select the value of $k$ that minimizes the average of $\CV(k)$
over all $K \times L$ folds (choosing the smallest value of $k$ in the event
of a tie).

@@

It will suffice to analyze a single fold of the cross-validation procedure.
As in in section~\ref{sec:gabriel-cv-algorithm} we assume that the $P$
variables of the data set have been partitioned into $p$ predictor variables
represented in vector~$X$ and $q$ response variables represented in
vector~$Y$.  The $N$ observations have been divided into two sets: $n$ train
observations and $m$ test observations.  The following theorem gives
conditions for Gabriel CV to recover the true number of clusters in the
absence of noise.

\begin{theorem}\label{thm:self-consistency}
Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
Suppose that
\begin{enumerate}[label=(\roman*)]

  \item There are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with the $g$th
    cluster center partitioned as $\mu(g) = \bigl(\muX(g), \muY(g)\bigr)$ with
    $\muX(g) \in \R^p$ and $\muY(g) \in \R^q$ for $g = 1, \dotsc, K$.

  \item The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct; likewise,
    $\muY(1), \dotsc,\muY(K)$ are all distinct.

  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.

  \item The training set contains at least one member of each cluster:
    for all $g$ in the range $1, \dotsc, K$, there exists at least one $i$ in
    the range $1, \dotsc, n$ such that $G_i = g$.

  \item \label{asn:self-consistency-test}
    The test set contains at least one member of each cluster: for all $g$ in
    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
    \dotsc, n+m$ such that $G_i = g$.

\end{enumerate}
Then $CV(k) < CV(K)$ for $k < K$, and $CV(k) = CV(K)$ for $k > K$, so that
Gabriel CV correctly chooses $k = K$.
\end{theorem}

@@

The proof of above theorem follows from following two lemmas:
\begin{lemma}
Suppose that the assumptions of Theorem $1$ are in force.  If $k < K$, then
$\CV(k) > 0$.
\end{lemma}
\begin{proof}
By definition,
\[
  \CV(k)
    =
      \sum_{i=n+1}^{n+m}
        \| Y_i - \bmuY (\hGX_i) \|^2,
\]
where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all $K$ distinct values
in $\{ \muY(1), \dotsc, \muY(K) \}$.  However, there are only $k$ distinct
values of $\bmuY(g)$.  Thus, at least one summand
\(
  \| Y_i - \bmuY(\hGX_i) \|^2
\)
is nonzero.  Therefore,
\(
  \CV(k) > 0.
\)
\end{proof}

\begin{lemma}
	If the assumptions of Theorem $1$ hold, when $k=K$, the CV Error $=0$
\end{lemma}
\begin{proof}
	It's sufficient to show that when $k=K$, $Error_k=0$. By assumption $3$ and $5$, training set contains $K$ centers. Which also implies $\mathbf{X}^R_{train}$ contains $K$ distinct $x^R_i$ $(\mu^R_1,...,\mu^R_K)$. When apply K-means on $\mathbf{X}^R_{train}$ with parameter $K$, the unique solution of 
	\[ \underset{\left(\hat{\mu}^R\left(g_1\right),...,\hat{\mu}^R\left(g_K\right)\right)}{\operatorname{argmin}} \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^R_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \]
	is 
\begin{equation}	
	\{\hat{\mu}^R \left(g_j\right) = \mu^R_j, j=1,...,K\}
\end{equation}	
with 
\begin{equation}	 
	\hat{g} \left(x^R_i\right) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
\end{equation}	
Where $\hat{g} \left(x^R_i\right) \in  \mathcal{G}$ is the assigned cluster of $x^R_i$ by K-means. One can easily check that such solution is unique and above objective function attains its lower bound $0$.\\\\
By Proposition $1$, $\mathcal{C}$ has re-substitution error $0$, which means its prediction error is $0$ on $\mathbf{X}^L_{test}$ as $\forall x^L_i \in \mathbf{X}^L_{test}, x^L_i \in \mathbf{X}^L_{train}$ by assumption $1,3,5$. Hence the predicted cluster $\hat{g} \left(x^L_i\right)$ for $x_i$ is the correct one $\hat{g} \left(x^R_i\right)$, which determined by $x^R_i$. By equation $(3)$, we have $x^R_i = \mu^R \left(c_i\right) = \mu^R_j$ if $\hat{g} \left(x^L_i\right) = \hat{g} \left(x^R_i\right) = g_j$. Therefore,
\[ Error_K = \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 =  \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - \mu^R_j||^2= 0\]
The last equation holds because of equation $(2)$.
\end{proof}
